---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mptstan

<!-- badges: start -->
<!-- badges: end -->

Estimate multinomial processing tree (MPT) models, a model class popular in cognitive psychology, in a Bayesian framework using package `brms` (i.e., ultimately `Stan`). This allows specifying models with any type of multivariate normal hierarchical or multilevel structure, including models with single and multiple random effects (i.e., crossed random effects). Model specification can be tailored for each model parameter. 

## Installation

You can install the development version of `mptstan` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mpt-network/mptstan")
```

## Example

We show the analysis of a recognition memory data set (from Singmann, Kellen, & Klauer, 2013) using the unsure-extended 2-high threshold model to a dataset investigating the other-race effect (i.e., a study with two different types of old and new items, own-race faces and other-race faces). This data is available in `mptstan` as `skk13`. We will analyse this data using crossed-random effects for participants and items.


```{r, message=FALSE}
library(mptstan)
library("bayestestR")
options(contrasts=c('contr.sum', 'contr.equalprior_pairs'))
str(skk13)
```

### Step 1: Create MPT Model Object

The first step when using `mptstan` is the creation of a MPT model object using `make_mpt()` (which creates an object of class `mpt_model`). 

`make_mpt()` can read MPT models in both the commonly used `EQN` model format (e.g., used by `TreeBUGS`) and the `easy` format introduced by `MPTinR`. 

```{r}
# For the easy EQN format, we just need the EQN file location:
EQNFILE <- system.file("extdata", "u2htm.eqn", package = "mptstan")
u2htsm_model <- make_mpt(EQNFILE) ## make_mpt() auto-detects EQN files from name
u2htsm_model

## Alternatively, we can just enter the equations and use the easy format.
u2htm <- "
# Old Items
Do + (1 - Do) * (1 - g1) * g2
(1 - Do) * g1
(1 - Do) * (1 - g1) * (1 - g2)

# New Items
(1 - Dn) * (1 - g1) * g2
(1 - Dn) * g1
Dn + (1 - Dn) * (1 - g1) * (1 - g2)
"
# for the easy format, we need to specify tree names and category names
u2htsm_model_2 <- make_mpt(text = u2htm, 
                           trees = c("old", "new"),
                           categories = rep(c("old", "unsure", "new"), 2))
u2htsm_model_2
```

### Step 2: Create Formula (Optional)

```{r}
u2htm_formula <- mpt_formula(resp ~ race + (race|s|id) + (1|p|stim), 
                             model = u2htsm_model)
```


## Step 3: Fit Model

```{r fit, cache=TRUE, results='hide', message=FALSE}
fit_skk <- mpt(u2htm_formula, data = skk13,
               tree = "type",
               cores = min(c(4, parallel::detectCores()))) ## uses multicore
```

```{r}
fit_skk
```


### Data Source

Singmann, H., Kellen, D., & Klauer, K. C. (2013). Investigating the Other-Race Effect of Germans towards Turks and Arabs using Multinomial Processing Tree Models. In M. Knauff, M. Pauen, N. Sebanz, & I. Wachsmuth (Eds.), Proceedings of the 35th Annual Conference of the Cognitive Science Society (pp. 1330–1335). Austin, TX: Cognitive Science Society.
http://singmann.org/download/publications/SKK-CogSci2013.pdf
