---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mptstan

<!-- badges: start -->
<!-- badges: end -->

Estimate multinomial processing tree (MPT) models, a model class popular in cognitive psychology, in a Bayesian framework using package `brms` (i.e., ultimately `Stan`). This allows specifying models with any type of multivariate normal hierarchical or multilevel structure, including models with single and multiple random effects (i.e., crossed random effects). Model specification can be tailored for each model parameter. 

## Installation

You can install the development version of `mptstan` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mpt-network/mptstan")
```

## Example

See `?mpt`

```{r example}
library(mptstan)
## basic example code
```
